Endogenous Macrodynamics in Algorithmic Recourse

### 2022-08-15 dev/paper/paper.tex
more work on results

\begin{document}
\maketitle
\begin{abstract}
Existing work on Counterfactual Explanations (CE) and Algorithmic
Recourse (AR) has largely been limited to the static setting and focused
on single individuals: given some estimated model the goal is to find
valid counterfactuals for individual instance that fulfill various
desiderata. The ability of such counterfactuals to handle dynamics like
data and model drift remains a largely unexplored research challenge at
this point. There has also been surprisingly little work on the related
question of how the actual implementation of recourse by one individual
may affect other individuals. Through this work we aim to close that gap
by systematizing and extending existing knowledge. We first show that
many of the existing methodologies can be collectively described by a
generalized framework. We then argue that the existing framework fails
to account for a hidden external cost of recourse, that only reveals
itself when studying the endogenous dynamics of recourse at the group
level. Through simulation experiments involving various popular
counterfactual generators and several benchmark datasets, we generate a
total XX million Counterfactual Explanations and study the resulting
domain and model shifts. We find that the induced shifts are substantial
enough to likely impede the applicability Algorithmic Recourse in
practice. Fortunately, we find various potential mitigation strategies
that can be used in combination with existing approaches. Our simulation
framework for studying recourse dynamics is fast and open-sourced.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, sharp corners, enhanced, interior hidden, borderline west={3pt}{0pt}{shadecolor}, frame hidden, breakable]}{\end{tcolorbox}}\fi

@@

\hypertarget{sec-empirical-2-mitigate}{%
\subsection{Potential Mitigation
Strategies}\label{sec-empirical-2-mitigate}}

@@

\hypertarget{more-conservative-decision-thresholds}{%
\subsubsection{More Conservative Decision
Thresholds}\label{more-conservative-decision-thresholds}}

@@

\hypertarget{classifier-preserving-roar}{%
\subsubsection{Classifier Preserving
ROAR}\label{classifier-preserving-roar}}

@@

\hypertarget{gravitational-counterfactual-explanations}{%
\subsubsection{Gravitational Counterfactual
Explanations}\label{gravitational-counterfactual-explanations}}

@@

Figure~\ref{fig-mitigation} shows an illustrative example that
demonstrates the differences in counterfactual outcomes when using the
various mitigation strategies compared to the baseline approach, that is
Wachter with \(\gamma=0.5\): choosing a higher decision threshold pushes
the counterfactual a little further into the target domain; this effect
is even stronger for ClapROAR; finally, using the Gravitational
generator the counterfactual ends up all the way inside the target
domain in the neighbourhood of \(\bar{x}\)\footnote{In order for the
  Graviational generator and ClapROAR to work as expected, one needs to
  ensure that counterfactual search continues, independent of the
  threshold probability \(\gamma\).}.

@@

\caption{\label{fig-mitigation}The differences in counterfactual
outcomes when using the various mitigation strategies compared to the
baseline approach, that is Wachter with \(\gamma=0.5\)}

@@

Our findings indicate that all three mitigation strategies are at least
at par with Latent Space generators with respect to their effectiveness
at mitigating domain and model shifts
(Figure~\ref{fig-mitigate-results}) \ldots{}

An interesting finding is also that the proposed strategies can have a
complementary effect when used in combination with Latent Space
generators. In experiments we conducted on the synthetic data, the
benefits of Latent Space generators were exacerbated further when using
a more conservative threshold or combining it with the penalties
underlying Gravitational and ClapROAR.

@@

@@

\caption{\label{fig-mitigate-results}PLACEHOLDER: Results for mitigating
strategies.}

>>>>

\begin{document}
\maketitle
\begin{abstract}
Existing work on Counterfactual Explanations (CE) and Algorithmic
Recourse (AR) has largely been limited to the static setting and focused
on single individuals: given some estimated model the goal is to find
valid counterfactuals for individual instance that fulfill various
desiderata. The ability of such counterfactuals to handle dynamics like
data and model drift remains a largely unexplored research challenge at
this point. There has also been surprisingly little work on the related
question of how the actual implementation of recourse by one individual
may affect other individuals. Through this work we aim to close that gap
by systematizing and extending existing knowledge. We first show that
many of the existing methodologies can be collectively described by a
generalized framework. We then argue that the existing framework fails
to account for a hidden external cost of recourse, that only reveals
itself when studying the endogenous dynamics of recourse at the group
level. Through simulation experiments involving various popular
counterfactual generators and several benchmark datasets, we generate a
total XX million Counterfactual Explanations and study the resulting
domain and model shifts. We find that the induced shifts are substantial
enough to likely impede the applicability Algorithmic Recourse in
practice. Fortunately, we find various potential mitigation strategies
that can be used in combination with existing approaches. Our simulation
framework for studying recourse dynamics is fast and open-sourced.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, boxrule=0pt, sharp corners, frame hidden, breakable, borderline west={3pt}{0pt}{shadecolor}, enhanced]}{\end{tcolorbox}}\fi

@@

\hypertarget{sec-empirical-2-mitigate}{%
\section{Potential Mitigation
Strategies}\label{sec-empirical-2-mitigate}}

@@

\hypertarget{more-conservative-decision-thresholds}{%
\subsection{More Conservative Decision
Thresholds}\label{more-conservative-decision-thresholds}}

@@

\hypertarget{classifier-preserving-roar}{%
\subsection{Classifier Preserving
ROAR}\label{classifier-preserving-roar}}

@@

\hypertarget{gravitational-counterfactual-explanations}{%
\subsection{Gravitational Counterfactual
Explanations}\label{gravitational-counterfactual-explanations}}

@@

Figure~\ref{fig-mitigation} hows an illustrative example that
demonstrates the differences in counterfactual outcomes when using the
various mitigation strategies compared to the baseline approach, that is
Wachter with \(\gamma=0.5\): choosing a higher decision threshold pushes
the counterfactual a little further into the target domain; this effect
is even stronger for ClapROAR; finally, using the Gravitational
generator the counterfactual ends up all the way inside the target
domain in the neighbourhood of \(\bar{x}\)\footnote{In order for the
  Graviational generator and ClapROAR to work as expected, one needs to
  ensure that counterfactual search continues, independent of the
  threshold probability \(\gamma\).}.

@@

\caption{\label{fig-mitigation}Illustrative example demonstrating the
properties of the various mitigation strategies.}

@@

Our findings indicate that all three mitigation strategies are at least
at par with Latent Space generators with respect to their effectiveness
at mitigating domain and model shifts. Figure~\ref{fig-mitigate-results}
presents a subset of the evaluation metrics for our synthetic data with
overlapping classes. The top row in Figure~\ref{fig-mitigate-results}
indicates that while domain shifts are of roughly the same magnitude for
both Wachter and Latent Space generators, our proposed strategies
effectively mitigate these shifts. ClapROAR appears to be particularly
effective, which is positively surprising, since it is designed to
explicitly address model shifts, not domain shifts. As evident from the
middle row in Figure~\ref{fig-mitigate-results} model shifts can also be
reduced: for the deep ensemble Latent Space search yields results that
are at par with the mitigation strategies, while for both the simple MLP
and logistic regression our simple strategies are actually more
effective. The same overall pattern can be observed for out-of-sample
model performance. With respect to the other synthetic datasets we found
the following: for the Moons dataset the emerging patterns are largely
the same, but the estimated model shifts are insignificant as noted
earlier; the same holds for the Circles dataset, but there is no
significant reduction in model performance for our neural networks; in
the case of linearly separable data we find the Gravitational generator
to be most effective at mitigating shifts.

@@

{\centering \includegraphics[width=0.45\textwidth,height=\textheight]{www/mitigation_synthetic_results.png}

@@

\caption{\label{fig-mitigate-results}The differences in counterfactual
outcomes when using the various mitigation strategies compared to the
baseline approach, that is Wachter with \(\gamma=0.5\). Results for
synthetic data with overlapping classes. The shown model MMD (PP MMD)
was computed over a meshgrid of points. Error bars indicate the standard
deviation across folds.}

An interesting finding is also that the proposed strategies can have a
complementary effect when used in combination with Latent Space
generators. In experiments we conducted on the synthetic data, the
benefits of Latent Space generators were exacerbated further when using
a more conservative threshold or combining it with the penalties
underlying Gravitational and ClapROAR. A snapshot of the results is
shown in Figure~\ref{fig-mitigate-latent-results}.

{\centering \includegraphics[width=9cm,height=9cm]{www/mitigation_synthetic_latent_results.png}

\caption{\label{fig-mitigate-latent-results}Combinining various
mitigation strategies with Latent Space search. Results for synthetic
data with overlapping classes. The shown model MMD (PP MMD) was computed
over a meshgrid of points. Error bars indicate the standard deviation
across folds.}
